---
title: "HW1: Sequoia NP Climate Trends"
author: "Nick McManus"
date: "2023-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(lubridate)
library(Kendall)
library(cowplot)
```


```{r}
### read in data
seq_clim <- read_csv("AshMt_GHCND_1927_2020.csv") %>% 
  janitor::clean_names()

### make sure date is correct class
as.Date(seq_clim$date)
# class(seq_clim$date)

```


### Running basic quality checks
Now that we've read in the data, the next step is quality checking! Let's make a plot to see what data are missing.

```{r dailyplots, echo=TRUE}
#### Max air temp -------------------------------------
## plot
ggplot(seq_clim, aes(date, tmax)) +
  geom_line() + 
  labs(y="Daily Maximum Temperature (degrees F)", x="Date")

## number of NAs
  # sum(is.na(seq_clim$tmax))
  # [1] 279
  # (sum(is.na(seq_clim$tmax)))/(count(seq_clim))
  # 0.008270342	


### Min air temp
## plot
ggplot(seq_clim, aes(date, tmin))+
  geom_line() + 
  labs(y="Daily Minimum Temperature (degrees F)", x="Date")

## number of NAS
  # sum(is.na(seq_clim$tmin))
  # [1] 322
  # (sum(is.na(seq_clim$tmin)))/(count(seq_clim))
  # 0.009544983


### Precip ---------------------------------------------
## plot
ggplot(seq_clim, aes(date, prcp)) +
  geom_line() + 
  labs(y="Daily Rainfall (in)", x="Date")

## number of NAS
  # sum(is.na(seq_clim$prcp))
  # [1] 1136
  # (sum(is.na(seq_clim$prcp)))/(count(seq_clim))
  # 0.03367423	
```


Not a crazy amount of NAs for temp, so going to fill them in with some assumptions
```{r}
### max temp
## values are either NA or not reasonable for that location (here, below 30F)
fillrow = which(is.na(seq_clim$tmax) | seq_clim$tmax <= 30)
fillrow = fillrow[2:length(fillrow)]
seq_clim$tmax[fillrow]=(seq_clim$tmax[fillrow+1]+seq_clim$tmax[fillrow-1])/2
ggplot(seq_clim, aes(date, seq_clim$tmax))+geom_line()+ labs(y="Daily Maximum Temperature (degrees F)", x="Date")

### min temp
## appears that there is at least one min temp value that is incorrect
fillrow = which(is.na(seq_clim$tmin) | seq_clim$tmin <= 15)
fillrow = fillrow[2:length(fillrow)]
seq_clim$tmin[fillrow]=(seq_clim$tmin[fillrow+1]+seq_clim$tmin[fillrow-1])/2
ggplot(seq_clim, aes(date, seq_clim$tmin))+geom_line()+ labs(y="Daily Maximum Temperature (degrees F)", x="Date")
```

Too many NAs for precip, and some span for large segments of the year. It *might* be safe to assume that these indicate no precipitation, but could be due to station functionality. For hot days, especially in summer, it is likely safe to assume that no precipitation occured. We'll fill in 0 for NA values in days above 70 degrees F, and leave the rest as NAs to be conservative. 
```{r}
fillrow = which((is.na(seq_clim$prcp)) & seq_clim$tmax >= 70)
## fill in data in the missing rows with zeros
seq_clim$prcp[fillrow]=0
## replot to make sure it works
ggplot(seq_clim, aes(date, prcp))+geom_line()+ labs(y="Daily rainfall (mm)", x="Date")

### Check the new number of NAs
  # sum(is.na(seq_clim$prcp))
  # [1] 276
```



### Looking for Trends

Now that the data has been cleaned up, it's time to actually do the trend analysis. BUT, trends can be 'swamped' by variation; in this case, the seasonal cycle is quite large. We also need to consider autocorrelation! 

So here let's try doing some aggregation to reduce the noise. As a simple example, we can try annual averages.

First, let's determine how many NAs are in each year to determine if it's responsible to use average values even with NAs.
```{r}
seq_clim_nas <- seq_clim %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(year) %>% 
  summarize(tmin_na = sum(is.na(tmin)),
            tmax_na = sum(is.na(tmax)))
```

Looks fairly spread out, so let's make sure to not remove a whole year due to a couple NAs for temp.
```{r annual, echo=TRUE}
### Find avg min and max temps and total precip by year
seq_clim_trend <- seq_clim %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(year) %>% 
  summarize(tmax_avg = mean(tmax, na.rm = T),
            tmin_avg = mean(tmin, na.rm = T),
            precip = sum(prcp, na.rm = T))

### Plot values for temp
a = ggplot(seq_clim_trend, aes(x=year, tmax_avg)) +
  geom_point(col='firebrick3', alpha = 0.5, size = 1.6) +
  scale_y_continuous(limits=c(min(seq_clim_trend$tmin_avg), max(seq_clim_trend$tmax_avg))) +
  geom_point(data=seq_clim_trend, aes(x=year, tmin_avg), 
             col="dodgerblue4", alpha = 0.5, size = 1.6) +
  theme_minimal() +
  labs(x = element_blank(),
       y = 'Temperature (F)') +
  theme(
    axis.title.y = element_text(vjust = 2, size = 11, face = 'bold'),
    axis.text.x = element_text(size = 11, face = 'bold'),
    axis.text.y = element_text(size = 11)
  )

# now lets add a trend line
a = a + stat_smooth(method="lm", col="firebrick")
final_temp <- a + stat_smooth(data=seq_clim_trend, aes(x=year,tmin_avg), col="dodgerblue4", method="lm")
final_temp
## save plot
ggsave(filename = 'avg_temps_sequoia.jpeg', plot = final_temp, width = 8, height = 5, units = "in")

### Plot values for precip
b = ggplot(seq_clim_trend, aes(x=year, precip)) +
  scale_y_continuous(limits=c(min(seq_clim_trend$precip), max(seq_clim_trend$precip))) +
  geom_point(data=seq_clim_trend, aes(x=year, precip), 
             col="darkorchid4", alpha = 0.5, size = 1.6) +
  theme_minimal() +
  labs(x = element_blank(),
       y = "Precipitation (in)") +
  theme(
    axis.title.y = element_text(vjust = 2, size = 11, face = 'bold'),
    axis.text.x = element_text(size = 11, face = 'bold'),
    axis.text.y = element_text(size = 11)
  )

precip_final <- b + stat_smooth(data = seq_clim_trend, aes(x=year, precip), col = 'darkorchid4', method = 'lm')
## save okit
ggsave(filename = 'avg_precip_sequoia.jpeg', plot = precip_final, width = 8, height = 5, units = "in")


### combine into one?
  # plot_grid(final_temp, precip_final,
  #           align = 'V',
  #           ncol = 1,
  #           labels = c('A', 'B'),
  #           hjust = -.5,
  #           vjust = 1)
```


Now let's calculate the slope for avg minimum temps:
```{r regressionline, echo=TRUE}


res=lm(tmin_avg~year, data=seq_clim_trend)
summary(res)
confint(res,"year", level=0.95)

ggplot(seq_clim_trend, aes(x=year, y=tmin_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")
```
Minimum temp is increasing by 0.0008 degrees F per year, but is not statistically significant (p = 0.89).

Let's do the same for avg maximum temps:
```{r}
res=lm(tmax_avg~year, data=seq_clim_trend)
summary(res)
confint(res,"year", level=0.95)

ggplot(seq_clim_trend, aes(x=year, y=tmax_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")
```
Max temps are increase at rate of 0.017 degrees per year, with significant p-value of 0.001

Finally, let's look at precip
```{r}
res=lm(precip~year, data=seq_clim_trend)
summary(res)
confint(res,"year", level=0.95)

ggplot(seq_clim_trend, aes(x=year, y=precip)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")
```
Looks like precipitation has decreased by -0.045 in/year, but not super significant (p-value: 0.181)


#### Compare decades

We might also cut the data into specific periods and see how the slope is changing as a function of time. We can see that some outliers more recently with the minimum temp data. 

```{r subset, echo=TRUE}
### TMIN split
## early portion
res_early=lm(tmin_avg~year, data=subset(seq_clim_trend, seq_clim_trend$year %in% c(1927:1960)))
summary(res_early)
confint(res_early,"dt", level=0.90)
ggplot(subset(seq_clim_trend, seq_clim_trend$year %in% c(1927:1960)), 
       aes(x=year, y=tmin_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) + 
  geom_smooth(method="lm")

# last two decades
res_late=lm(tmin_avg~year, data=subset(seq_clim_trend, seq_clim_trend$year %in% c(2000:2020)))
summary(res_late)
confint(res_late,"dt", level=0.90)
ggplot(subset(seq_clim_trend, seq_clim_trend$year %in% c(2000:2020)), 
       aes(x=year, y=tmin_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")

```

Looking at the last two decades, minimum temps have increased by 0.16 F/year, but p-value only 0.097. In earlier period, decreased temperature (-0.025F/year) with a p-value of 0.335.

Let's see if splitting up the max temp does anything. 
```{r}
### TMAX split
## early portion
res_early=lm(tmax_avg~year, data=subset(seq_clim_trend, seq_clim_trend$year %in% c(1927:1960)))
summary(res_early)
confint(res_early,"dt", level=0.90)
ggplot(subset(seq_clim_trend, seq_clim_trend$year %in% c(1927:1960)), aes(x=year, y=tmax_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) + 
  geom_smooth(method="lm")

# last four decades
res_late=lm(tmax_avg~year, data=subset(seq_clim_trend, seq_clim_trend$year %in% c(1980:2020)))
summary(res_late)
confint(res_late,"dt", level=0.90)
ggplot(subset(seq_clim_trend, seq_clim_trend$year %in% c(1980:2020)), aes(x=year, y=tmax_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")

```

Here we see that the earlier trend was slightly increasing (0.016F/year) with a p-value of 0.482, while in the recent couple decades we've seen warming (0.026F/year) with a p-value of 0.156. 

Finally, let's see how precip differed
```{r}
### PRECIP split
## early portion
res_early=lm(precip~year, data=subset(seq_clim_trend, seq_clim_trend$year %in% c(1927:1950)))
summary(res_early)
confint(res_early,"dt", level=0.90)
ggplot(subset(seq_clim_trend, seq_clim_trend$year %in% c(1927:1950)), aes(x=year, y=precip)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) + 
  geom_smooth(method="lm")

# last two decades
res_late=lm(precip~year, data=subset(seq_clim_trend, seq_clim_trend$year %in% c(2000:2020)))
summary(res_late)
confint(res_late,"dt", level=0.90)
ggplot(subset(seq_clim_trend, seq_clim_trend$year %in% c(2000:2020)), aes(x=year, y=precip)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")
```
In the earlier decades, annual precip changed by 0.11 in/year, but not significantly (p-value: 0.652)

In recent decades, annual precip has decreased (-0.457in/year) but not quite significantly (p-value = 0.0968)




### Stats tests

Regression assumes a linear relationship - and normally distributed data - sometimes that isn't true, we can use non-parameteric tests to look for trends. In these cases, the Mann-Kendall test is commonly used.

tau ranges from -1 to 1 and denotes the "strength" of the trend; p-value denotes significance. Strength however can not be interpreted as slope!

```{r kendall, echo=TRUE}

MannKendall(seq_clim_trend$tmin_avg)
MannKendall(seq_clim_trend$tmax_avg)
MannKendall(seq_clim_trend$precip)

```

TMIN: tau = 0.0597, 2-sided pvalue =0.39575
TMAX: tau = 0.224, 2-sided pvalue =0.0013853
PRECIP: tau = -0.0888, 2-sided pvalue =0.20621

we can see that tmax has the strongest trend and is the most significant. Precipitation changes have the next strongest trend, though it's not quite as significant. 


We'll see if the temperatures differ between early/late time periods significantly. We can do this with both a t-test, which assumes normal distribution, and a Mann-Whitney Wilcoxon test, which does not. 

```{r}
### TMIN
t.test(subset(seq_clim_trend$tmin_avg, seq_clim_trend$year %in% 1927:1950), 
       subset(seq_clim_trend$tmin_avg, seq_clim_trend$year %in% 2000:2020))

wilcox.test(subset(seq_clim_trend$tmin_avg, seq_clim_trend$year %in% 1927:1950), 
            subset(seq_clim_trend$tmin_avg, seq_clim_trend$year %in% 2000:2020))
```

Doesn't seem significant enough. Let's look at max temperature
```{r}
### TMAX
t.test(subset(seq_clim_trend$tmax_avg, seq_clim_trend$year %in% 1927:1950), 
       subset(seq_clim_trend$tmax_avg, seq_clim_trend$year %in% 2000:2020))

wilcox.test(subset(seq_clim_trend$tmax_avg, seq_clim_trend$year %in% 1927:1950), 
            subset(seq_clim_trend$tmax_avg, seq_clim_trend$year %in% 2000:2020))
```
These show there is a significant difference for both test, with a slightly lower p-value for the Wilcoxon test (0.005) than the t-test (0.008).

Finally, let's compare the precip.
```{r}
### PRECIP
t.test(subset(seq_clim_trend$precip, seq_clim_trend$year %in% 1927:1950), 
       subset(seq_clim_trend$precip, seq_clim_trend$year %in% 2000:2020))

wilcox.test(subset(seq_clim_trend$precip, seq_clim_trend$year %in% 1927:1950), 
            subset(seq_clim_trend$precip, seq_clim_trend$year %in% 2000:2020))
```
Here we see that the difference is fairly significant according to the t-test (p-value = 0.055) and some-what less significant according to the rank-sum test (p-value = 0.099)



### Aggregation: seasonal!

An alternative approach to aggregation (mean by year)
is to look at a particular season, lets say we want to look only at summer (July and August)

Min temp:
```{r alternative, echo=TRUE}
## create a variable
seq_clim_season <- seq_clim %>% 
  mutate(season = ifelse(month(date) %in% c(12,1,2), 1, ifelse(month(date) %in% c(3:5),2, ifelse(month(date) %in% c(6:8),3,4))))

clim_byseason = seq_clim_season %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(year, season) %>% 
  summarize(tmax_avg=mean(tmax), 
            tmin_avg=mean(tmin), 
            precip=sum(prcp))

## look only at summer
clim_summer = subset(clim_byseason, clim_byseason$season==3)

```

```{r}
### TMIN

## plot
ggplot(clim_summer, aes(x=year, y=tmin_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4) +
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")+labs(y=" Summer Minimum Daily Temperature C")

res=lm(tmin_avg~year, data=clim_summer)
summary(res)
confint(res,"wy", level=0.95)

```


Increase of 0.012F/year in the summer, but not significant (p-value = 0.192)

Let's look at max temp:
```{r}
### TMAX

##plot 
summer_tmax <- ggplot(clim_summer, aes(x=year, y=tmax_avg)) + 
  stat_summary(fun="mean", geom="point", col="firebrick", size=1.6, alpha = 0.5) +
  theme_minimal() +
  geom_smooth(method="lm", col = 'firebrick4')+
  labs(y = "Temperature (F)",
       x = element_blank()) +
  theme(
    axis.title.y = element_text(vjust = 2, size = 11, face = 'bold'),
    axis.text.x = element_text(size = 11, face = 'bold'),
    axis.text.y = element_text(size = 11)
  )

res=lm(tmax_avg~year, data=clim_summer)
summary(res)
confint(res,"wy", level=0.95)

## save plot
ggsave(filename = 'summer_tmax_sequoia.jpeg', plot = summer_tmax, width = 8, height = 5, units = "in")

```

Summer max temps have significantly increased 0.0218 F/year (p-value = 0.004)

Now let's see how precipitation have changed over winters 
```{r}
## look only at winter
clim_winter = subset(clim_byseason, clim_byseason$season==1)

##plot 
winter_precip <- ggplot(clim_winter, aes(x=year, y=precip)) + 
  stat_summary(fun="mean", geom="point", col="darkorchid3", size=1.6, alpha = 0.5) +
   geom_smooth(method="lm", col = 'darkorchid4')+
   labs(y = "Precipitation (in)",
       x = element_blank()) +
  theme_minimal()+
   theme(
    axis.title.y = element_text(vjust = 2, size = 11, face = 'bold'),
    axis.text.x = element_text(size = 11, face = 'bold'),
    axis.text.y = element_text(size = 11)
  )

res=lm(precip~year, data=clim_winter)
summary(res)
confint(res,"wy", level=0.95)

## save plot
ggsave(filename = 'winter_precip_sequoia.jpeg', plot = winter_precip, width = 8, height = 5, units = "in")

```
Not a significant change (p-value = 0.838) of -0.006F/year. 


Let's look at old/new differences by season

```{r}
### TMAX split
## early portion
res_early=lm(tmax_avg~year, data=subset(clim_summer, clim_summer$year %in% c(1927:1950)))
summary(res_early)
confint(res_early,"dt", level=0.90)
ggplot(subset(clim_summer, clim_summer$year %in% c(1927:1950)), aes(x=year, y=tmax_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) + 
  geom_smooth(method="lm")

# last four decades
res_late=lm(tmax_avg~year, data=subset(clim_summer, clim_summer$year %in% c(2000:2020)))
summary(res_late)
confint(res_late,"dt", level=0.90)
ggplot(subset(clim_summer, clim_summer$year %in% c(2000:2020)), aes(x=year, y=tmax_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")


```
Early time non-significantly decreased slightly. Newer time shows increase in average summer temps by 0.0671F/year (p-value = 0.011)

```{r}
### TMAX
t.test(subset(clim_summer$tmax_avg, clim_summer$year %in% 1927:1950), 
       subset(clim_summer$tmax_avg, clim_summer$year %in% 2000:2020))

wilcox.test(subset(clim_summer$tmax_avg, clim_summer$year %in% 1927:1950), 
            subset(clim_summer$tmax_avg, clim_summer$year %in% 2000:2020))
```

We can also see the difference between these two time periods is significantly different by t-test (p = 0.019) and rank-sum (p = 0.033).


Now let's check for winter precipitation.
```{r}
### precip split
## early portion
res_early=lm(precip~year, data=subset(clim_winter, clim_winter$year %in% c(1927:1950)))
summary(res_early)
confint(res_early,"dt", level=0.90)
ggplot(subset(clim_winter, clim_winter$year %in% c(1927:1950)), aes(x=year, y=precip)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) + 
  geom_smooth(method="lm")

# last four decades
res_late=lm(precip~year, data=subset(clim_winter, clim_winter$year %in% c(2000:2020)))
summary(res_late)
confint(res_late,"dt", level=0.90)
ggplot(subset(clim_winter, clim_winter$year %in% c(2000:2020)), aes(x=year, y=precip)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")


```

Here we see a non-significant change in average winter precipitation levels by year for the early period. The later decades are also not sig (p-value = 0.19) for a -0.13in/year decline.

Although these trends don't appear significant, maybe the difference in means are.

```{r}
### precip
t.test(subset(clim_winter$precip, clim_winter$year %in% 1927:1950), 
       subset(clim_winter$precip, clim_winter$year %in% 2000:2020))

wilcox.test(subset(clim_winter$precip, clim_winter$year %in% 1927:1950), 
            subset(clim_winter$precip, clim_winter$year %in% 2000:2020))
```
No significant difference in means.


