---
title: "HW1: Chico Climate Trends"
author: "Nick McManus"
date: "2023-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(lubridate)
library(Kendall)
```


```{r}
### read in data
chico_clim <- read_csv("Chico_GHCND_3301914.csv") %>% 
  janitor::clean_names()
head(chico_clim)

### make sure date is correct class
as.Date(chico_clim$date)
# class(chico_clim$date)

```


### Running basic quality checks
Now that we've read in the data, the next step is quality checking! Let's make a plot to see what data are missing.

```{r dailyplots, echo=TRUE}
#### Max air temp
## plot
ggplot(chico_clim, aes(date, tmax)) +
  geom_line() + 
  labs(y="Daily Maximum Temperature (degrees F)", x="Date")
## number of NAs
# sum(is.na(chico_clim$tmax))
# [1] 2448
# (sum(is.na(chico_clim$tmax)))/(count(chico_clim))
# 0.06086827


### Min air temp
## plot
ggplot(chico_clim, aes(date, tmin))+
  geom_line() + 
  labs(y="Daily Minimum Temperature (degrees F)", x="Date")

## number of NAS
# sum(is.na(chico_clim$tmin))
# [1] 2478
# (sum(is.na(chico_clim$tmin)))/(count(chico_clim))
# 0.0616142


### Precip
## plot
ggplot(chico_clim, aes(date, prcp)) +
  geom_line() + 
  labs(y="Daily Rainfall (in)", x="Date")

## number of NAS
# sum(is.na(chico_clim$prcp))
# [1] 940
# (sum(is.na(chico_clim$prcp)))/(count(chico_clim))
# 0.02337262	
```


Not a crazy amount of NAs, so going to fill them in with some assumptions
```{r}
### first deal with mising precip
fillrow = which(is.na(chico_clim$prcp))
## fill in data in the missing rows with zeros
chico_clim$prcp[fillrow]=0
## replot to make sure it works
ggplot(chico_clim, aes(date, prcp))+geom_line()+ labs(y="Daily rainfall (mm)", x="Date")


### now max temp
## values are either NA or not reasonable for that location (here, below 30F)
fillrow = which(is.na(chico_clim$tmax) | chico_clim$tmax < 30)
fillrow = fillrow[2:length(fillrow)]
chico_clim$tmax[fillrow]=(chico_clim$tmax[fillrow+1]+chico_clim$tmax[fillrow-1])/2
ggplot(chico_clim, aes(date, chico_clim$tmax))+geom_line()+ labs(y="Daily Maximum Temperature (degrees F)", x="Date")
```

### Looking for Trends

Now that the data has been cleaned up, it's time to actually do the trend analysis. BUT, trends can be 'swamped' by variation; in this case, the seasonal cycle is quite large. We also need to consider autocorrelation! 

So here let's try doing some aggregation to reduce the noise. As a simple example, we can try annual averages.

```{r annual, echo=TRUE}
### Find avg min and max temps and total precip by year

### remove 2023 data bc incomplete for year
chico_clim_mwy <- chico_clim %>% 
  mutate(year = lubridate::year(date)) %>% 
  filter(!year %in% c(2023)) %>% 
### Find avg min and max temps and total precip by year
  group_by(year) %>% 
  summarize(tmax_avg = mean(tmax, na.rm = F),
            tmin_avg = mean(tmin, na.rm = F),
            precip = sum(prcp))

### Plot values
ggplot(chico_clim_mwy, aes(x = year, y = tmax_avg)) +
  geom_point(col="red") +
  scale_y_continuous(limits = c(min(chico_clim_mwy$tmin_avg), max(chico_clim_mwy$tmax_avg))) +
  geom_point(data = chico_clim_mwy, aes(x = year, tmin_avg), col="blue")



a = ggplot(chico_clim_mwy, aes(x=year, tmax_avg)) +
  geom_point(col="red") +
  scale_y_continuous(limits=c(min(chico_clim_mwy$tmin_avg), max(chico_clim_mwy$tmax_avg))) +
  geom_point(data=chico_clim_mwy, aes(x=year, tmin_avg), col="blue")

# now lets add a trend line
a = a + stat_smooth(method="lm", col="red")
a
a + stat_smooth(data=chico_clim_mwy, aes(x=year,tmin_avg), col="blue", method="lm")

```

```{r wy, echo=TRUE}

# now lets add a trend line
a =a+stat_smooth(method="lm", col="red")
a
a+ stat_smooth(data=clim.mwy, aes(x=dt,tmin), col="blue", method="lm")

```


Now let's calculate the slope (or how quickly temperatures are rising; we do this with linear regression)

```{r regressionline, echo=TRUE}


res=lm(tmin_avg~year, data=chico_clim_mwy)
summary(res)
confint(res,"year", level=0.95)
ggplot(chico_clim_mwy, aes(x=year, y=tmin_avg)) + stat_summary(fun.y="mean", geom="point", col="red", size=4)+theme(axis.text=element_text(size=14, face="bold"), axis.title=element_text(size=14, face="bold")) + geom_smooth(method="lm")
```










