---
title: "HW1: Sequoia NP Climate Trends"
author: "Nick McManus"
date: "2023-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(lubridate)
library(Kendall)
```


```{r}
### read in data
seq_clim <- read_csv("AshMt_GHCND_1927_2020.csv") %>% 
  janitor::clean_names()

### make sure date is correct class
as.Date(seq_clim$date)
# class(seq_clim$date)

```


### Running basic quality checks
Now that we've read in the data, the next step is quality checking! Let's make a plot to see what data are missing.

```{r dailyplots, echo=TRUE}
#### Max air temp -------------------------------------
## plot
ggplot(seq_clim, aes(date, tmax)) +
  geom_line() + 
  labs(y="Daily Maximum Temperature (degrees F)", x="Date")

## number of NAs
  # sum(is.na(seq_clim$tmax))
  # [1] 279
  # (sum(is.na(seq_clim$tmax)))/(count(seq_clim))
  # 0.008270342	


### Min air temp
## plot
ggplot(seq_clim, aes(date, tmin))+
  geom_line() + 
  labs(y="Daily Minimum Temperature (degrees F)", x="Date")

## number of NAS
  # sum(is.na(seq_clim$tmin))
  # [1] 322
  # (sum(is.na(seq_clim$tmin)))/(count(seq_clim))
  # 0.009544983


### Precip ---------------------------------------------
## plot
ggplot(seq_clim, aes(date, prcp)) +
  geom_line() + 
  labs(y="Daily Rainfall (in)", x="Date")

## number of NAS
  # sum(is.na(seq_clim$prcp))
  # [1] 1136
  # (sum(is.na(seq_clim$prcp)))/(count(seq_clim))
  # 0.03367423	
```


Not a crazy amount of NAs for temp, so going to fill them in with some assumptions
```{r}
### max temp
## values are either NA or not reasonable for that location (here, below 30F)
fillrow = which(is.na(seq_clim$tmax) | seq_clim$tmax <= 30)
fillrow = fillrow[2:length(fillrow)]
seq_clim$tmax[fillrow]=(seq_clim$tmax[fillrow+1]+seq_clim$tmax[fillrow-1])/2
ggplot(seq_clim, aes(date, seq_clim$tmax))+geom_line()+ labs(y="Daily Maximum Temperature (degrees F)", x="Date")

### min temp
## appears that there is at least one min temp value that is incorrect
fillrow = which(is.na(seq_clim$tmin) | seq_clim$tmin <= 15)
fillrow = fillrow[2:length(fillrow)]
seq_clim$tmin[fillrow]=(seq_clim$tmin[fillrow+1]+seq_clim$tmin[fillrow-1])/2
ggplot(seq_clim, aes(date, seq_clim$tmin))+geom_line()+ labs(y="Daily Maximum Temperature (degrees F)", x="Date")
```

Too many NAs for precip, and some span for large segments of the year. It *might* be safe to assume that these indicate no precipitation, but could be due to station functionality. For hot days, especially in summer, it is likely safe to assume that no precipitation occured. We'll fill in 0 for NA values in days above 70 degrees F, and leave the rest as NAs to be conservative. 
```{r}
fillrow = which((is.na(seq_clim$prcp)) & seq_clim$tmax >= 70)
## fill in data in the missing rows with zeros
seq_clim$prcp[fillrow]=0
## replot to make sure it works
ggplot(seq_clim, aes(date, prcp))+geom_line()+ labs(y="Daily rainfall (mm)", x="Date")

### Check the new number of NAs
  # sum(is.na(seq_clim$prcp))
  # [1] 276
```



### Looking for Trends

Now that the data has been cleaned up, it's time to actually do the trend analysis. BUT, trends can be 'swamped' by variation; in this case, the seasonal cycle is quite large. We also need to consider autocorrelation! 

So here let's try doing some aggregation to reduce the noise. As a simple example, we can try annual averages.

First, let's determine how many NAs are in each year to determine if it's responsible to use average values even with NAs.
```{r}
seq_clim_nas <- seq_clim %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(year) %>% 
  summarize(tmin_na = sum(is.na(tmin)),
            tmax_na = sum(is.na(tmax)))
```

Looks fairly spread out, so let's make sure to not remove a whole year due to a couple NAs for temp.
```{r annual, echo=TRUE}
### Find avg min and max temps and total precip by year
seq_clim_trend <- seq_clim %>% 
  mutate(year = lubridate::year(date)) %>% 
  group_by(year) %>% 
  summarize(tmax_avg = mean(tmax, na.rm = T),
            tmin_avg = mean(tmin, na.rm = T),
            precip = sum(prcp))

### Plot values
ggplot(seq_clim_trend, aes(x = year, y = tmax_avg)) +
  geom_point(col="red") +
  scale_y_continuous(limits = c(min(seq_clim_trend$tmin_avg), max(seq_clim_trend$tmax_avg))) +
  geom_point(data = seq_clim_trend, aes(x = year, tmin_avg), col="blue")



a = ggplot(seq_clim_trend, aes(x=year, tmax_avg)) +
  geom_point(col="red") +
  scale_y_continuous(limits=c(min(seq_clim_trend$tmin_avg), max(seq_clim_trend$tmax_avg))) +
  geom_point(data=seq_clim_trend, aes(x=year, tmin_avg), col="blue")

# now lets add a trendd line
a = a + stat_smooth(method="lm", col="red")
a
a + stat_smooth(data=seq_clim_trend, aes(x=year,tmin_avg), col="blue", method="lm")

```


Now let's calculate the slope for avg minimum temps:
```{r regressionline, echo=TRUE}


res=lm(tmin_avg~year, data=seq_clim_trend)
summary(res)
confint(res,"year", level=0.95)

ggplot(seq_clim_trend, aes(x=year, y=tmin_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")
```
Minimum temp is increasing by 0.0008 degrees F per year, but is not statistically significant (p = 0.89).

Let's do the same for avg maximum temps:
```{r}
res=lm(tmax_avg~year, data=seq_clim_trend)
summary(res)
confint(res,"year", level=0.95)

ggplot(seq_clim_trend, aes(x=year, y=tmax_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")
```
Max temps are increase at rate of 0.017 degrees per year, with significant p-value of 0.001


We might also cut the data into specific periods and see how the slope is changing as a function of time. We can see that some outliers more recently with the minimum temp data. 

```{r subset, echo=TRUE}
### TMIN split
## early portion
res_early=lm(tmin_avg~year, data=subset(seq_clim_trend, seq_clim_trend$year %in% c(1927:1980)))
summary(res_early)
confint(res_early,"dt", level=0.90)
ggplot(subset(seq_clim_trend, seq_clim_trend$year %in% c(1927:1980)), aes(x=year, y=tmin_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) + 
  geom_smooth(method="lm")

# last two decades
res_late=lm(tmin_avg~year, data=subset(seq_clim_trend, seq_clim_trend$year %in% c(2001:2015)))
summary(res_late)
confint(res_late,"dt", level=0.90)
ggplot(subset(seq_clim_trend, seq_clim_trend$year %in% c(2001:2020)), aes(x=year, y=tmin_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")


```

Looking at the last two decades, minimum temps have increased by 0.16 F/year, but p-value only 0.121. In earlier period, decreased temperature (-0.012F/year) with a p-value of 0.292.

Let's see if splitting up the max temp does anything. 
```{r}
### TMAX split
## early portion
res_early=lm(tmax_avg~year, data=subset(seq_clim_trend, seq_clim_trend$year %in% c(1927:1950)))
summary(res_early)
confint(res_early,"dt", level=0.90)
ggplot(subset(seq_clim_trend, seq_clim_trend$year %in% c(1927:1950)), aes(x=year, y=tmax_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), 
        axis.title=element_text(size=14, face="bold")) + 
  geom_smooth(method="lm")

# last three decades
res_late=lm(tmax_avg~year, data=subset(seq_clim_trend, seq_clim_trend$year %in% c(1975:2020)))
summary(res_late)
confint(res_late,"dt", level=0.90)
ggplot(subset(seq_clim_trend, seq_clim_trend$year %in% c(1975:2020)), aes(x=year, y=tmax_avg)) + 
  stat_summary(fun="mean", geom="point", col="red", size=4)+
  theme(axis.text=element_text(size=14, face="bold"), axis.title=element_text(size=14, face="bold")) +
  geom_smooth(method="lm")


```

Here we see that the earlier trend was slightly cooling (-0.06F/year) with a p-value of 0.085, while in the recent couple decades we've seen warming (0.036F/year) with a p-value of 0.02. 


