---
title: "HW2"
author: "Nick McManus"
date: "2023-04-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

### Read in necessary packages
library(lubridate)
library(ggplot2)
library(tidyverse)
library(chron)
library(ncdf4)  ##netCDF files
library(RColorBrewer)
library(lattice)
library(abind)
```



## Read in data

```{r}
# path and filename for data
dname <- "ts"  # this is the name of the variable you want to look at

### Create names for all file paths
# GISS Historical
gissHist1 <- "data/ts_Amon_GISS-E2-1-G_historical_r2i1p1f2_gn_185001-190012.nc"  
gissHist2 <- "data/ts_Amon_GISS-E2-1-G_historical_r2i1p1f2_gn_190101-195012.nc"  
gissHist3 <- "data/ts_Amon_GISS-E2-1-G_historical_r2i1p1f2_gn_195101-200012.nc" 
gissHist4 <- "data/ts_Amon_GISS-E2-1-G_historical_r2i1p1f2_gn_200101-201412.nc" 

# GISS SSP3
giss_ssp3_1 <- "data/ts_Amon_GISS-E2-1-G_ssp370_r1i1p3f1_gn_201501-205012.nc"
giss_ssp3_2 <- "data/ts_Amon_GISS-E2-1-G_ssp370_r1i1p3f1_gn_205101-210012.nc"

# GISS SSP5
giss_ssp5_1 <- "data/ts_Amon_GISS-E2-1-G_ssp585_r3i1p5f1_gn_201501-205012.nc"
giss_ssp5_2 <- "data/ts_Amon_GISS-E2-1-G_ssp585_r3i1p5f1_gn_205101-210012.nc"


### Read in data
# GISS Historical
gissHist1 <- nc_open(gissHist1)
gissHist2 <- nc_open(gissHist2)
gissHist3 <- nc_open(gissHist3)
gissHist4 <- nc_open(gissHist4)

# GISS ssp3
giss_ssp3_1 <- nc_open(giss_ssp3_1)
giss_ssp3_2 <- nc_open(giss_ssp3_2)

# GISS ssp5
giss_ssp5_1 <- nc_open(giss_ssp5_1)
giss_ssp5_2 <- nc_open(giss_ssp5_2)
```


## Time stuff

```{r}
### Adjust the time for each file
# Historical
timeH1 <- ncvar_get(gissHist1,"time")
tunitsH1 <- ncatt_get(gissHist1,"time","units")
timeH2 <- ncvar_get(gissHist2,"time")
tunitsH2 <- ncatt_get(gissHist2,"time","units")
timeH3 <- ncvar_get(gissHist3,"time")
tunitsH3 <- ncatt_get(gissHist3,"time","units")
timeH4 <- ncvar_get(gissHist4,"time")
tunitsH4 <- ncatt_get(gissHist4,"time","units")

# SSP3
time3_1 <- ncvar_get(giss_ssp3_1,"time")
tunits3_1 <- ncatt_get(giss_ssp3_1,"time","units")
time3_2 <- ncvar_get(giss_ssp3_2,"time")
tunits3_2 <- ncatt_get(giss_ssp3_2,"time","units")

# SSP5
time5_1 <- ncvar_get(giss_ssp5_1,"time")
tunits5_1 <- ncatt_get(giss_ssp5_1,"time","units")
time5_2 <- ncvar_get(giss_ssp5_2,"time")
tunits5_2 <- ncatt_get(giss_ssp5_2,"time","units")


# Concatenate the historical data with the two SSP scenario datasets

time_history = c(timeH1,timeH2, timeH3, timeH4)
time_ssp3 = c(time3_1, time3_2)
time_total = c(timeH1,timeH2, timeH3, timeH4, time3_1, time3_2)


rtime=seq.Date(as.Date("1850/01/01"),as.Date("2100/12/01"),by="months")


# Do the weird time split thing that I think you only need to do once, but im confused about this
tustr <- strsplit(tunitsH1$value, " ")
tdstr <- strsplit(unlist(tustr)[3], "-")
tmonth <- as.integer(unlist(tdstr)[2])
tday <- as.integer(unlist(tdstr)[3])
tyear <- as.integer(unlist(tdstr)[1])

# Follow Sam's code to do this weird time thing for both datasets
rtime_ssp3 <- chron(time_ssp3,origin=c(tmonth, tday, tyear))
rtime_ssp5 <- chron(time_ssp5,origin=c(tmonth, tday, tyear))
```


## Surface temp

```{r}
### Pull out the surface temperature for each
# Historical
tsH1 <- ncvar_get(gissHist1, "ts")
tsH2 <- ncvar_get(gissHist2, "ts")
tsH3 <- ncvar_get(gissHist3, "ts")
tsH4 <- ncvar_get(gissHist4, "ts")

# SSP3
ts_3_1 <- ncvar_get(giss_ssp3_1, "ts")
ts_3_2 <- ncvar_get(giss_ssp3_2, "ts")

# SSP5
ts_5_1 <- ncvar_get(giss_ssp5_1, "ts")
ts_5_2 <- ncvar_get(giss_ssp5_2, "ts")


### Bind the TS together for each scenario
ts_ssp3=abind(tsH1,tsH2, tsH3, tsH4, ts_3_1, ts_3_2)
ts_ssp5=abind(tsH1,tsH3, tsH3, tsH4, ts_5_1, ts_5_2)
```

## Lat/lon

```{r}
# Fix the lat and lon
lon <- ncvar_get(gissHist1,"lon")
nlon <- dim(lon)
lat <- ncvar_get(gissHist1,"lat")
nlat <- dim(lat)

# Choose your location (these are for the amazon)
lats=which(lat >= -20 & lat <= 5)
lons=which(lon >= 45 & lon <= 75)
```


# Read in the data for graphing 

```{r}
# Average the data (i had to mess around with the numeric)
tsavg_ssp3 <- apply(ts_ssp3[lons,lats,],3,mean)
tsavg_ssp5 <- apply(ts_ssp5[lons,lats,],3,mean)


# Make it a graphable dataframe
clim_ssp3 <- data.frame(time=time_total, tsavg = tsavg_ssp3)
clim_ssp5 <- data.frame(time = time_total, tsavg = tsavg_ssp5)


# aggregate temp data by annual averages
yrclim_giss_ssp3 = clim_ssp3 %>% 
  group_by(year(rtime)) %>%
  summarize(Tann = mean(tsavg))
yrclim_giss_ssp3$dt = unique(year(rtime))

yrclim_giss_ssp5 = clim_ssp5 %>% 
  group_by(year(rtime)) %>% 
  summarize(Tann=mean(tsavg))
yrclim_giss_ssp5$dt = unique(year(rtime))

# Read in MIROC model
yrclim_miroc_ssp3 = read_csv("data/miroc_ssp3_ts_1850_2100.csv")
yrclim_miroc_ssp5 = read_csv("data/miroc_ssp5_ts_1850_2100.csv")



# Subset the historical data
giss_clim_hist <- yrclim_giss_ssp3 %>% 
  filter(dt <= 2014)
miroc_clim_hist <- yrclim_miroc_ssp3 %>% 
  filter(dt <= 2014)

# Subset the SSP3 data
giss_clim_ssp3 <- yrclim_giss_ssp3 %>% 
  filter(dt > 2014)
miroc_clim_ssp3 <- yrclim_miroc_ssp3 %>% 
  filter(dt > 2014)

# Subset the SSP5 data
giss_clim_ssp5 <- yrclim_giss_ssp5 %>% 
  filter(dt > 2014)
miroc_clim_ssp5 <- yrclim_miroc_ssp5 %>% 
  filter(dt > 2014)


# run linear regression so I can compare the slopes of the data
giss_clim_hist_lm <- lm((Tann-273.15)~dt,data=giss_clim_hist)
giss_clim_ssp3_lm <- lm((Tann-273.15)~dt,data=giss_clim_ssp3)
giss_clim_ssp5_lm <- lm((Tann-273.15)~dt,data=giss_clim_ssp5)

miroc_clim_hist_lm <- lm((Tann-273.15)~dt,data=giss_clim_hist)
miroc_clim_ssp3_lm <- lm((Tann-273.15)~dt,data=giss_clim_ssp3)
miroc_clim_ssp5_lm <- lm((Tann-273.15)~dt,data=giss_clim_ssp5)
```

Wiggly plot
```{r}
ggplot() +
  geom_line(data = yrclim_miroc_ssp3, aes(x = dt, y = Tann - 273.15, color = "red")) +
  geom_line(data = yrclim_miroc_ssp5, aes(x = dt, y = Tann - 273.15, color = 'blue')) +
  labs(y = "Concordia Station Temperature (degrees C)", x = "Year", title = "Mean Surface Air Temperature Change in Concordia Station, Antartica", subtitle = "Models Used: NASA GISS-E2.1-G and NASA GISS-E2.1-H", caption = "Smoothing Method Employed: Linear Regression", color = "Legend") + 
  geom_smooth(method = "lm") +
  geom_vline(xintercept = 2014, linetype = "dashed") + 
  scale_color_manual(values = c('blue', 'cadetblue3', 'darkgoldenrod1', 'chocolate1')) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5), 
        legend.position = "right", 
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.key.size = unit(1.5, "lines"),
        legend.margin = margin(t = 0, b = 0, r = 10, l = 10),
        legend.box.just = "right")
```

Smooth plot
```{r}
ggplot() +
  theme_minimal() +
  labs(y="Temperature (C)\n", 
       x="\nYear", 
       title = "Mean Annual Surface Temperature Trends for Amazon Biome") + 
  theme(plot.title = element_text(hjust=0.3),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10)) +
  xlim(1850, 2100) +
  ylim(27, 32.5) +
  geom_smooth(data=yrclim_giss_ssp3, aes(dt, Tann-273.15),col = '#2274A5',fill="#2274A5", size =1) +
  geom_smooth(data=yrclim_giss_ssp5, aes(dt, Tann-273.15), col = '#efa947',fill="#efa947", size = 1) +
  geom_vline(xintercept=2014,linetype="dashed") +
  annotate("text", label = "Historical Model\n1850-2015", x = 1950, y = 29, size = 4, colour = "black", family = "sans") +
  annotate("text", label = "SSP5-8.5", x = 2075, y = 32, size = 4, colour = "#efa947", family = "sans") +
  annotate("text", label = "SSP3-7.0", x = 2090, y = 29.5, size = 4, colour = "#2274A5", family = "sans")
```































## Read in data

```{r}
# path and filename for data
ncpath <- "data/"   # path (directory)
ncname <- "ts_Amon_GISS-E2-1-G_historical_r2i1p1f2_gn_185001-190012.nc"  # CESM1 filename
ncfname <- paste(ncpath, ncname, sep="")
dname <- "ts"  # this is the name of the variable you want to look at

ncin <- nc_open(ncfname)
print(ncin)
```


Merge historical/SSP3 data for NASA GISS

```{r}
# path and filename for data
ncpath <- "data/"   # path (directory)
dname <- "ts"  # this is the name of the variable you want to look at

ncname1 <- "ts_Amon_GISS-E2-1-G_historical_r2i1p1f2_gn_185001-190012.nc"
ncname2 <- "ts_Amon_GISS-E2-1-G_historical_r2i1p1f2_gn_190101-195012.nc"
ncname3 <- "ts_Amon_GISS-E2-1-G_historical_r2i1p1f2_gn_195101-200012.nc"
ncname4 <- "ts_Amon_GISS-E2-1-G_historical_r2i1p1f2_gn_200101-201412.nc"
ncname5 <- "ts_Amon_GISS-E2-1-G_ssp370_r1i1p3f1_gn_201501-205012.nc"
ncname6 <- "ts_Amon_GISS-E2-1-G_ssp370_r1i1p3f1_gn_205101-210012.nc"

ncfname1 <- paste(ncpath, ncname1, sep="")
ncfname2 <- paste(ncpath, ncname2, sep="")
ncfname3 <- paste(ncpath, ncname3, sep="")
ncfname4 <- paste(ncpath, ncname4, sep="")
ncfname5 <- paste(ncpath, ncname5, sep="")
ncfname6 <- paste(ncpath, ncname6, sep="")

ncin_giss1 <- nc_open(ncfname1)
ncin_giss2 <- nc_open(ncfname2)
ncin_giss3 <- nc_open(ncfname3)
ncin_giss4 <- nc_open(ncfname4)
ncin_giss5 <- nc_open(ncfname5)
ncin_giss6 <- nc_open(ncfname6)
# print(ncin_giss1)

# Get tas information from all files
tas1 <- ncvar_get(ncin_giss1, "ts")
tas2 <- ncvar_get(ncin_giss2, "ts")
tas3 <- ncvar_get(ncin_giss3, "ts")
tas4 <- ncvar_get(ncin_giss4, "ts")
tas5 <- ncvar_get(ncin_giss5, "ts")
tas6 <- ncvar_get(ncin_giss6, "ts")


# Concatenate tas information along the time dimension (dimension 3)
tas=abind(tas1,tas2,tas3,tas4,tas5,tas6, along=3)
```

```{r e3smtime_ssp}
rtime_giss=seq.Date(as.Date("1850/01/01"),as.Date("2100/12/01"),by="months")

```


The following commands read in the longitude and latitude information, and store the lengths of each axis in variables 'nlon' and 'nlat'.

```{r readcoords}
lon <- ncvar_get(ncin_giss1,"lon")
nlon <- dim(lon)
lat <- ncvar_get(ncin_giss1,"lat")
nlat <- dim(lat)

head(lat)
head(lon)

```


Now we can redo the lat/lon regional extraction:
```{r getreg_e3sm}
lats=which(lat >= 32 & lat <= 35)
lons=which(lon >= 241 & lon <= 243)


tsavg_giss <- apply(tas[lons,lats,],3,mean)
```



Now once again, we read in the time coordinate information. We need to concatenate the time information from the files to get a time series of the full time period:

```{r}
time1 <- ncvar_get(ncin_giss1,"time")
tunits1 <- ncatt_get(ncin_giss1,"time","units")
time2 <- ncvar_get(ncin_giss2,"time")
tunits2 <- ncatt_get(ncin_giss2,"time","units")
time3 <- ncvar_get(ncin_giss3,"time")
tunits3 <- ncatt_get(ncin_giss3,"time","units")
time4 <- ncvar_get(ncin_giss4,"time")
tunits4 <- ncatt_get(ncin_giss4,"time","units")

time=c(time1,time2,time3,time4)
print(tunits4)
```
Use chron package to split up the "tunits" values (which is a string) into year, month, and day

```{r}
tustr <- strsplit(tunits1$value, " ")
tdstr <- strsplit(unlist(tustr)[3], "-")
tmonth <- as.integer(unlist(tdstr)[2])
tday <- as.integer(unlist(tdstr)[3])
tyear <- as.integer(unlist(tdstr)[1])
rtime <- chron(time1,origin=c(tmonth, tday, tyear))
```









